---
title: "leapR Paper Examples"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: |
  %\VignetteIndexEntry{leapR}   
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```



# Load libraries needed
```{r libraries, message=FALSE, warning = FALSE}
# load the core libraries
library(leapR)
library(gplots)
library(rmarkdown)
# plotting helpers used in this vignette
library(ggplot2)
library(dplyr)
library(tibble)
library(stringr)
library(BiocFileCache)
```

#  Example data
A sample data set is included that is from the CPTAC study of 169 ovarian
tumors. We include the dataset as a \code{SummarizedExperiment} object, 
containing three assays (transcriptomics, global proteomics, and 
phosphoproteomics) to enable interoperability with other tools, and 
store example file as `rda` on 
[Figshare](https://figshare.com/account/projects/251807/articles/29250845)
as example.

This data can be loaded as follows:
```{r omicsdata, message=FALSE, warning=FALSE}

#currently using the BiocFileCache though i'm not sure it helps
path <- tools::R_user_dir("leapR", which = "cache")
bfc <- BiocFileCache(path, ask = FALSE)

url <- "https://api.figshare.com/v2/file/download/56536217"
pc <- bfcadd(bfc, "pdat", fpath = url)
load(pc)

url <- "https://api.figshare.com/v2/file/download/56536214"
tc <- bfcadd(bfc, "tdat", fpath = url)
load(tc)

url <- "https://api.figshare.com/v2/file/download/56536211"
phc <- bfcadd(bfc, "phdat", fpath = url)
load(phc)

```

We also have local data we can load

```{r load lists}

data(shortlist)
data(longlist)
data(ncipid)
data(kinasesubstrates)
## columns that we want to use for results

cols_to_display <- c("ingroup_n", "outgroup_n", "background_n", 
                     "pvalue", "BH_pvalue")

```

# Figure 2
We will compare the ability of transcriptomics, proteomics, and 
phosphoproteomics to inform about
differences between short and long surviving patient groups. In addition
to other methods, we also employ a `calcTTest` function that takes 
two sets of samples from the `SummarizedExperiment` object and computes 
the t-test between them. The results are then stored in the `rowData` of the 
same object, so that they can be used for enrichment later on. 

This spans the multiple enrichment methods in leapR and also includes 
multi-omics. 

The resulting heatmap is presented as Figure 2 in the paper.

``` {r figure_2, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
# load the single omic and multi-omic pathway databases
data("krbpaths")
data("mo_krbpaths")

# comparison enrichment in transcriptional data
transdata.comp.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_comparison",
  eset = tset,
  assay_name = "transcriptomics",
  primary_columns = shortlist,
  secondary_columns = longlist
)

# comparison enrichment in proteomics data
# this is the same code used above, just repeated here for clarity
protdata.comp.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_comparison",
  eset = pset,
  assay_name = "proteomics",
  primary_columns = shortlist,
  secondary_columns = longlist
)

# comparison enrichment in phosphoproteomics data
phosphodata.comp.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_comparison",
  eset = phset,
  assay_name = "phosphoproteomics",
  primary_columns = shortlist,
  secondary_columns = longlist, id_column = "hgnc_id"
)


# set enrichment in transcriptomics data
# perform the comparison t-test
tset <- leapR::calcTTest(tset, assay_name = "transcriptomics", 
                         shortlist, longlist)


## now we need to run enrichment in sets with target list, not eset
transdata.set.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  eset = tset,
  assay_name = "transcriptomics",
  enrichment_method = "enrichment_in_sets",
  primary_columns = "pvalue",
  greaterthan = FALSE, threshold = 0.05
)


pset <- leapR::calcTTest(pset, assay_name = "proteomics",
                         shortlist, longlist)

protdata.set.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  eset = pset,
  assay_name = "proteomics",
  enrichment_method = "enrichment_in_sets",
  primary_columns = "pvalue",
  greaterthan = FALSE, threshold = 0.05
)


phset <- leapR::calcTTest(phset, assay_name = "phosphoproteomics", 
                          shortlist, longlist)

phosphodata.set.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_sets",
  id_column = "hgnc_id",
  assay_name = "phosphoproteomics",
  eset = phset, primary_columns = "pvalue",
  greaterthan = FALSE, threshold = 0.05
)

# order enrichment in transcriptomics data
transdata.order.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_order",
  eset = tset,
  assay_name = "transcriptomics",
  primary_columns = "difference"
)

# order enrichment in proteomics data
protdata.order.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_order",
  eset = pset,
  assay_name = "proteomics",
  primary_columns = "difference"
)

# order enrichment in phosphoproteomics data


phosphodata.order.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_order",
  id_column = "hgnc_id",
  method = 'ztest', 
  eset = phset,
  assay_name = "phosphoproteomics",
  primary_columns = "difference"
)

# correlation difference in transcriptomics data
transdata.corr.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "correlation_comparison",
  eset = tset,
  assay_name = "transcriptomics",
  primary_columns = shortlist,
  secondary_columns = longlist
)
# correlation difference in proteomics data
protdata.corr.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "correlation_comparison",
  eset = pset,
  assay_name = "proteomics",
  primary_columns = shortlist,
  secondary_columns = longlist
)
# correlation difference in phosphoproteomics data
phosphodata.corr.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "correlation_comparison",
  eset = phset,
  assay_name = "phosphoproteomics",
  primary_columns = shortlist,
  secondary_columns = longlist, id_column = "hgnc_id"
)

# combine the omics data into one with prefix tags
comboset <- leapR::combine_omics(list(pset, phset, tset), 
                                 c(NA, "hgnc_id", NA))

# comparison enrichment for combodata
# when we use expression set, we do not need to use the mo_krbpaths 
#since the  id mapping column is used
combodata.enrichment.svl <- leapR::leapR(
  geneset = krbpaths, # mo_krbpaths,
  enrichment_method = "enrichment_comparison",
  eset = comboset,
  assay_name = "combined",
  primary_columns = shortlist,
  secondary_columns = longlist, id_column = "id"
)


# set enrichment in combo data
# perform the comparison t test
comboset <- leapR::calcTTest(comboset,
                             assay_name = "combined", 
                             shortlist, longlist)

combodata.set.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_sets",
  eset = comboset, primary_columns = "pvalue",
  assay_name = "combined",
  id_column = "id",
  greaterthan = FALSE, threshold = 0.05
)

# order enrichment in combo data
combodata.order.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "enrichment_in_order",
  assay_name = "combined",
  eset = comboset, primary_columns = "difference",
  id_column = "id"
)

# correlation difference in combo data
combodata.corr.enrichment.svl <- leapR::leapR(
  geneset = krbpaths,
  enrichment_method = "correlation_comparison",
  eset = comboset,
  assay_name = "combined",
  primary_columns = shortlist,
  id_column = "id",
  secondary_columns = longlist
)


# now take all these results and combine them into one figure
all_results <- list(
  transdata.comp.enrichment.svl,
  protdata.comp.enrichment.svl,
  phosphodata.comp.enrichment.svl,
  combodata.enrichment.svl,
  transdata.set.enrichment.svl,
  protdata.set.enrichment.svl,
  phosphodata.set.enrichment.svl,
  combodata.set.enrichment.svl,
  transdata.order.enrichment.svl,
  protdata.order.enrichment.svl,
  phosphodata.order.enrichment.svl,
  combodata.order.enrichment.svl,
  transdata.corr.enrichment.svl,
  protdata.corr.enrichment.svl,
  phosphodata.corr.enrichment.svl,
  combodata.corr.enrichment.svl
)

pathways_of_interest <- c(
  "KEGG_APOPTOSIS",
  "KEGG_CELL_CYCLE",
  "KEGG_ERBB_SIGNALING_PATHWAY",
  "KEGG_FOCAL_ADHESION",
  "KEGG_INSULIN_SIGNALING_PATHWAY",
  "KEGG_MAPK_SIGNALING_PATHWAY",
  "KEGG_MISMATCH_REPAIR",
  "KEGG_MTOR_SIGNALING_PATHWAY",
  "KEGG_OXIDATIVE_PHOSPHORYLATION",
  "KEGG_P53_SIGNALING_PATHWAY",
  "KEGG_PATHWAYS_IN_CANCER",
  "KEGG_PROTEASOME",
  "KEGG_RIBOSOME",
  "KEGG_VEGF_SIGNALING_PATHWAY",
  "KEGG_WNT_SIGNALING_PATHWAY"
)


results.frame <- data.frame(
  pathway = pathways_of_interest,
  td.comp = all_results[[1]][pathways_of_interest, "BH_pvalue"] < 0.05,
  pd.comp = all_results[[2]][pathways_of_interest, "BH_pvalue"] < 0.05,
  fd.comp = all_results[[3]][pathways_of_interest, "BH_pvalue"] < 0.05,
  cd.comp = all_results[[4]][pathways_of_interest, "BH_pvalue"] < 0.05,
  td.set = all_results[[5]][pathways_of_interest, "BH_pvalue"] < 0.05,
  pd.set = all_results[[6]][pathways_of_interest, "BH_pvalue"] < 0.05,
  fd.set = all_results[[7]][pathways_of_interest, "BH_pvalue"] < 0.05,
  cd.set = all_results[[8]][pathways_of_interest, "BH_pvalue"] < 0.05,
  td.order = all_results[[9]][pathways_of_interest, "BH_pvalue"] < 0.05,
  pd.order = all_results[[10]][pathways_of_interest, "BH_pvalue"] < 0.05,
  fd.order = all_results[[11]][pathways_of_interest, "BH_pvalue"] < 0.05,
  cd.order = all_results[[12]][pathways_of_interest, "BH_pvalue"] < 0.05,
  td.corr = all_results[[13]][pathways_of_interest, "BH_pvalue"] < 0.05,
  pd.corr = all_results[[14]][pathways_of_interest, "BH_pvalue"] < 0.05,
  fd.corr = all_results[[15]][pathways_of_interest, "BH_pvalue"] < 0.05,
  cd.corr = all_results[[16]][pathways_of_interest, "BH_pvalue"] < 0.05
)

results.frame.or <- data.frame(
  pathway = pathways_of_interest,
  td.comp = all_results[[1]][pathways_of_interest, "oddsratio"],
  pd.comp = all_results[[2]][pathways_of_interest, "oddsratio"],
  fd.comp = all_results[[3]][pathways_of_interest, "oddsratio"],
  cd.comp = all_results[[4]][pathways_of_interest, "oddsratio"],
  td.set = log(all_results[[5]][pathways_of_interest, "oddsratio"], 2),
  pd.set = log(all_results[[6]][pathways_of_interest, "oddsratio"], 2),
  fd.set = log(all_results[[7]][pathways_of_interest, "oddsratio"], 2),
  cd.set = log(all_results[[8]][pathways_of_interest, "oddsratio"], 2),
  td.order = all_results[[9]][pathways_of_interest, "oddsratio"],
  pd.order = all_results[[10]][pathways_of_interest, "oddsratio"],
  fd.order = all_results[[11]][pathways_of_interest, "oddsratio"],
  cd.order = all_results[[12]][pathways_of_interest, "oddsratio"],
  td.corr = all_results[[13]][pathways_of_interest, "oddsratio"],
  pd.corr = all_results[[14]][pathways_of_interest, "oddsratio"],
  fd.corr = all_results[[15]][pathways_of_interest, "oddsratio"],
  cd.corr = all_results[[16]][pathways_of_interest, "oddsratio"]
)

rownames(results.frame) <- results.frame[, 1]
rownames(results.frame.or) <- results.frame.or[, 1]
results.frame.sig <- results.frame[, 2:17] * results.frame.or[, 2:17]

heatmap.2(as.matrix(results.frame.sig[, c(1:4, 9:16)]), Colv = NA, 
          trace = "none", breaks = c(-1, -.1, -0.0001, 0, 0.1, 1), 
          col = c("blue", "lightblue", "grey", "pink", "red"), 
          dendrogram = "none")
```


Figure 3. An application of the KSEA-like approach in leapR as applied to our
          example data. In this example we are looking for known substrate sets
          of kinases (from Phosphosite Plus) that are enriched in the short vs
          long comparison of phosphopeptides.
``` {r figure_3, warning=FALSE, message=FALSE}
# this comparison of abundance in substrates between case and control
#     is lopsided in the sense that phosphorylation levels were previously
#     reported to be overall higher in the short survivors. Thus the
#     results are not terribly interesting (all kinases are in the same 
# direction)
phosphodata.ksea.comp.svl <- leapR::leapR(
  geneset = kinasesubstrates,
  enrichment_method = "enrichment_comparison",
  eset = phset,
  assay_name = "phosphoproteomics",
  primary_columns = shortlist, secondary_columns = longlist
)


# thus for the example we'll look at correlation between known substrates 
#   in the   case v control conditions
phosphodata.ksea.corr.svl <- leapR::leapR(
  geneset = kinasesubstrates,
  enrichment_method = "correlation_comparison",
  eset = phset,
  assay_name = "phosphoproteomics",
  primary_columns = shortlist,
  secondary_columns = longlist
)

# for the example we are using an UNCORRECTED PVALUE
#     which will allow us to plot more values, but
#     for real applications it's necessary to use the
#     CORRECTED PVALUE

# here are all the kinases *significant (*uncorrected) from the analysis
or <- order(phosphodata.ksea.corr.svl[, "pvalue"])
ksea_result <- phosphodata.ksea.corr.svl[or, ][1:9, ]
ksea_cols <- rep("grey", 9)
ksea_cols[which(ksea_result[, "oddsratio"] > 0)] <- "black"

# plot left panel: correlation significance of top most significant kinases
barplot(ksea_result[, "oddsratio"],
  horiz = TRUE, xlim = c(-1, 0.5),
  names.arg = rownames(ksea_result), las = 1, col = ksea_cols
)

# plot right panel: abundance comparison results of the same kinases
barplot(phosphodata.ksea.comp.svl[rownames(ksea_result), "oddsratio"],
  horiz = TRUE, names.arg = rownames(ksea_result), las = 1, col = "black"
)
```